---
title: "Pro-Life Attitude Consistency Across Political Issues"
author: "Kaya Vadhan"
description: "GOV50 Final Project 2023"
output:
  distill::distill_article:
    self_contained: false
---

## Background

Much pro-life rhetoric in the abortion debate surrounds the idea that abortion is the murder of an unborn child. Broadly, I would like to examine whether this pro-life abortion attitude is truly value-based, or whether it is specific to abortion and may be influenced or caused by other factors (such as partisanship). In this study, I will specifically look at the relationship between anti-abortion beliefs and beliefs on other political issues relating to murder. Opposition to the death-penalty should also fall under a pro-life attitude. Favoring bans on semi-automatic assault rifles, which have contributed to extreme numbers of deaths and shootings, should also fall under a pro-life attitude. Lastly, opposing an eagerness to use military force in international conflict should fall under an pro-life attitude, considering the thousands of civilian deaths resulting from an arguably excessive use of military force. Converse opinions on these topics would seemingly conflict with a pro-life value. By assessing consistency in pro-life sentiments across abortion, ideally, we will gain insight into whether being pro-life is a true value of individuals, or solely an abortion rhetoric. 

Hypotheses

I hypothesize that these attitudes will not be consistent because anti-death penalty, pro-gun control, and anti-military force attitudes are traditionally Democratic ideologies, whereas pro-life abortion attitudes are often Republican ideologies, suggesting some form of inconsistency. If this hypothesis is true, this would bring into question the underlying foundation of the pro-life anti-abortion argument and suggest that it is just a rhetoric, platform, or caused by other factors such as partisanship. As such, I will also be measuring the correlation between partisan self-identification and abortion attitudes. As abortion is seen as a very partisan issue, I do believe that there will be a strong correlation here.   

## Data

My data is cross-sectional, from the American Election Studies Data Center 2020 survey of 8280 individuals. The survey results for the variables I am looking at were done on a Likert scale. 

OUTCOME VARIABLE AND MEASUREMENT

Opposition to Abortion Rights Reduction: "Pro Abortion Rights Score" (anti pro-life attitude)

Survey question: Would you be pleased, upset, or neither pleased nor upset if the Supreme Court reduced abortion rights?

1. Extremely pleased
2. Moderately pleased
3. A little pleased
4. Neither pleased nor upset
5. A little upset
6. Moderately upset
7. Extremely upset

```{r}
library(tidyverse)
library(ggplot2)
dataset <- read_csv("anes2020.csv") |>
  select(V201342x, V201345x, V202344x, V201350, V201200) |>
  rename(
    abortion = V201342x,
    deathpenalty = V201345x,
    guncontrol = V202344x,
    militaryforce = V201350,
    id = V201200
  )
dataset <- dataset |>
  filter(deathpenalty > 0, guncontrol > 0, abortion > 0, militaryforce > 0, id > 0, id < 10)
head(dataset)
 
dataset |>
  ggplot(mapping = aes(x = abortion)) +
  geom_bar() +
  labs(
x = "Pro Abortion Rights Score",
y = "Respondent Count",
title = "Respondent Perspective on Abortion Rights"
)
```

Out of the thousands of respondents, from this graph, wee can see that the majority of respondents would be extremely upset at the reduction of abortion rights or are indifferent on the matter, so they are generally not "pro-lifers". However, there are still over 1000 respondents that fall on the pro-life half of the spectrum. 


EXPLANATORY VARIABLES AND MEASUREMENT

1) Opposition to the Death Penalty: "Anti Death Penalty Score" (pro-life attitude)

Survey question: Do you favor or oppose the death penalty for persons convicted of murder?

1. Favor strongly
2. Favor not strongly
3. Oppose not strongly
4. Oppose strongly

2) Opposition to banning assault rifles: "Pro Assault Rifles Score" (anti pro-life attitude)

Survey question: Do you favor, oppose, or neither favor nor oppose banning the sale of semi-automatic “assault-style” rifles?

1. Favor a great deal
2. Favor a moderate amount
3. Favor a little
4. Neither favor nor oppose
5. Oppose a little
6. Oppose a moderate amount
7. Oppose a great deal

3) Opposition to using military force in international conflict: "Anti Military Force Score" (pro-life attitude)

Survey question: How willing should the United States be to use military force to solve international problems? 

1. Extremely willing
2. Very willing
3. Moderately willing
4. A little willing
5. Not at all willing

4) Liberal-Conservative Self Identification: "Conservative Identification Score"

Survey question: Where would you place yourself on this scale [7PT SCALE LIBERAL-CONSERVATIVE]?

1. Extremely liberal
2. Liberal
3. Slightly liberal
4. Moderate; middle of the road
5. Slightly conservative
6. Conservative
7. Extremely conservative

Since opposition to the death penalty and opposition to using military force in international conflict are both pro-life attitudes, and opposition to abortion rights reduction is an anti pro-life attitude, a significant positive correlation between these variables would support my hypothesis that pro-life attitudes are not consistent between abortion and other issues. Since opposition to banning assault rifles is an anti pro-life attitude, just like opposition to abortion rights reduction, a significant negative correlation here would support my hypothesis that pro-life attitudes are not consistent between abortion and other issues. A significant negative correlation between degree of conservative identification and opposition of abortion right reduction would further support my hypothesis that the pro-life argument for abortion is not based in values, but instead a partisan identification or political argument. However, if there were significant correlations found in the other direction, this would suggest that pro-life attitudes are consistent among many political issues, and that the pro-life abortion argument is a true, consistent value, which I believe would give it more merit, as opposed to just political rhetoric.

## Results

```{r}
library(modelsummary)

ggplot(dataset, aes(x = deathpenalty, y = abortion)) +
geom_jitter() + geom_smooth(method = "lm", se = FALSE) +
labs(
x = "Anti Death Penalty Scale",
y = "Pro Abortion Rights Scale",
title = "Abortion Attitudes vs Death Penalty Attitudes"
)

ggplot(dataset, aes(x = militaryforce, y = abortion)) +
geom_jitter() + geom_smooth(method = "lm", se = FALSE) +
labs(
x = "Anti Military Force Scale",
y = "Pro Abortion Rights Scale",
title = "Abortion Attitudes vs Military Force Attitudes"
)

ggplot(dataset, aes(x = guncontrol, y = abortion)) +
geom_jitter() + geom_smooth(method = "lm", se = FALSE) +
labs(
x = "Pro Assault Rifle Scale",
y = "Pro Abortion Rights Scale",
title = "Abortion Attitudes vs Gun Control Attitudes"
)

regression <- lm(abortion ~ deathpenalty + militaryforce + guncontrol, data = dataset)
summary(regression)
```
Add description


Liberal-Conservative Identification

```{r}
##id
ggplot(dataset, aes(x = id, y = abortion)) +
geom_jitter() + geom_smooth(method = "lm", se = FALSE) +
labs(
x = "Conservative Identification Score",
y = "Pro Abortion Rights Scale",
title = "Abortion Attitudes vs Liberal-Conservative Identification"
)
idregression <- lm(abortion ~ id, data = dataset)
summary(idregression)
```
Description

Regression Summary

```{r}
var_labels <- c(
    "(Intercept)" = "Intercept",
    "abortion" = "Pro Abortion Rights Score",
    "deathpenalty" = "Anti Death Penalty Score",
    "guncontrol" = "Pro Assault Rifle Score",
    "militaryforce" = "Anti Military Force Score",
    "id" = "Conservative Identification Score"
  )
modelsummary::modelsummary(list("Pro-Life Attitudes" = regression, "Conservative Identification" = idregression),
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           coef_map = var_labels,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))

```

The scatterplots indicate noticable associations between abortion attitudes and death penalty attitudes, abortion attitudes and military force attitudes, abortion attitudes and gun control attitudes, and an extremely strong association between abortion attitudes and liberal-conservative identification. The association between pro abortion rights scores and anti death penalty scores, as well as and anti military force scores, are both positive, whereas the association with pro assault rifles scores and conservative identification scores are negative. 

After running a regression (not including conservative identification), we found the intercept to be 4.304, with a coefficient of .337 on the anti death penalty score, -.255 on the pro assault rifle score, and .151 on the anti military force score. This means that a 1 unit increase on the pro abortion rights scale is associated with an average .337 unit increase on the anti death penalty scale, an average .255 unit decrease on the pro assault rifle scale, and an average .121 unit increase on the anti military force scale. In other words, increasing abortion rights support is associated with increasing death penalty opposition, increasing gun control support, and increasing military force opposition. 

When running a separate regression to see whether ideological identification was related to abortion attitudes, the coefficient of -.798 means that every unit increase on the pro abortion rights scale is associated with an average .798 unit decrease on the liberal-conservative identification scale, indicating that as one identifies as more liberal, they are less likely to support abortion rights

Each of these correlations is statistically significant at the .01 level with the p-value for each being less than .001. However, we should not interpret this data causally, as there are many confounding factors that could contribute to one's beliefs on any of these topics or party identification, even if there is a causal relationship directionality is unclear, there is no clear reason indiciating that one of these factors has caused another, and lastly, we are not testing to examine causality, but instead consistency of attitudes, which can be measured with correlation. 

## Conclusion
Each of the explanatory variables we tested showed extremely statistically significant correlations with abortion attitudes. As I hypothesized, each case, the "pro-life" attitude on each topic (pro banning of assault rifles, anti-death penalty, anti-use of military force) is inversely associated with a "pro-life" abortion attitude (anti abortion). Further, pro-life abortion attitudes were strongly correlated with high levels of conservatism, as expected. This study provides significant evidence to support the hypothesis that the anti-abortion "pro-life" argument is not a true internal value, but could be simply political propaganda and rhetoric. However, there are still limitations to this inference. Treating Likert-scale results as a continuous variable and running a linear regression on them is not the most precise or ideal, and if we learned how to run other forms of regression and more statistical tests in this class, those may have been better to be used. Further, individuals may have pro-life values but multiple competing values that arise in these cases of the death penalty (in this case, the person being killed has already committed heinous crimes), use of international military force (in this case, there may be threat to the individual's life themselves), and gun control (in this case, there is the frequently discussed competing constitutional right to bear arms of the second amendment). So, it may not be that being "pro-life" is solely abortion rhetoric, but that in other cases, other values end up being weighed as more important. Lastly, extrapolating data from one survey with only a few questions asked about each of these issues can make it hard to gain actual perspective on the attitudes of respondents. Issues like question framing may create biases in our results. With more time and money, ideally, surveys would dive deep into each of these issues, asking similar questions in multiple different forms and asking specific nuanced questions that help tease out the underlying reasoning behind each of their attitudes. Overall, though, this study provides a great evidence for continued research on this topic. 