---
title: "Pro-Life Attitude Consistency Across Political Issues"
author: "Kaya Vadhan"
description: "GOV50 Final Project 2023"
output:
  distill::distill_article:
    self_contained: false
---

## Background

Background

Much pro-life rhetoric in the abortion debate surrounds the idea that abortion is the murder of an unborn child. Broadly, I would like to examine whether this pro-life abortion attitude is truly value-based, or whether it is specific to abortion and may be influenced or caused by other factors (such as partisanship). In this study, I will specifically look at the relationship between anti-abortion beliefs and beliefs on other political issues relating to murder. Opposition to the death-penalty should also fall under a pro-life attitude. Favoring bans on semi-automatic assault rifles, which have contributed to extreme numbers of deaths and shootings, should also fall under a pro-life attitude. Lastly, opposing an eagerness to use military force in international conflict should fall under an pro-life attitude, considering the thousands of civilian deaths resulting from an arguably excessive use of military force. Converse opinions on these topics would seemingly conflict with a pro-life value. By assessing consistency in pro-life sentiments across abortion, ideally, we will gain insight into whether being pro-life is a true value of individuals, or solely an abortion rhetoric. 

Hypotheses

I hypothesize that these attitudes will not be consistent because anti-death penalty, pro-gun control, and anti-military force attitudes are traditionally Democratic ideologies, whereas pro-life abortion attitudes are often Republican ideologies, suggesting some form of inconsistency. If this hypothesis is true, this would bring into question the underlying foundation of the pro-life anti-abortion argument and suggest that it is just a rhetoric, platform, or caused by other factors such as partisanship. As such, I will also be measuring the correlation between partisan self-identification and abortion attitudes. As abortion is seen as a very partisan issue, I do believe that there will be a strong correlation here.   

## Data

Data

My data is from the American Election Studies Data Center. The specific sample is from a 2020  data collection of 8280 individuals. The survey results for the variables I am looking at were done on a Likert scale. 

Outcome variable and measurement: Favoring of abortion rights (anti-pro-life attitude)

Survey question: Would you be pleased, upset, or neither pleased nor upset if the Supreme Court reduced abortion rights? [pro-choice]
1. Extremely pleased
2. Moderately pleased
3. A little pleased
4. Neither pleased nor upset
5. A little upset
6. Moderately upset
7. Extremely upset

```{r}
library(tidyverse)
library(ggplot2)
dataset <- read_csv("anes2020.csv") |>
  select(V201342x, V201345x, V202344x, V201350, V201200) |>
  rename(
    abortion = V201342x,
    deathpenalty = V201345x,
    guncontrol = V202344x,
    militaryforce = V201350,
    ideology = V201200
  )
dataset <- dataset |>
  filter(deathpenalty > 0, guncontrol > 0, abortion > 0, militaryforce > 0, ideology > 0, ideology < 10)
head(dataset)
 
dataset |>
  ggplot(mapping = aes(x = abortion)) +
  geom_bar() +
  labs(
x = "Degree of Upsetness at a Potential Reduction of Abortion Rights",
y = "Respondent Count",
title = "Respondent Perspective on Abortion Rights"
)
```

[EDITS HERE EDITS HERE EDITS HERE] 
It is important to note that scoring highly on the outcome variable would imply a pro-choice attitude (anti-pro-life attitude), not pro-life attitude. For that reason, when analyzing the data _____. Further, we can see that the majority of respondents would be extremely upset at the reduction of abortion rights or are indifferent on the matter. 


Explanatory variables and measurement:

1) Opposition to the Death Penalty (pro-life attitude)

Survey question: Do you favor or oppose the death penalty for persons convicted of murder?
1. Favor strongly
2. Favor not strongly
3. Oppose not strongly
4. Oppose strongly

2) Opposition to banning assault rifles (anti-pro-life attitude)

Survey question: Do you favor, oppose, or neither favor nor oppose banning the sale of semi-automatic “assault-style” rifles?

1. Favor a great deal
2. Favor a moderate amount
3. Favor a little
4. Neither favor nor oppose
5. Oppose a little
6. Oppose a moderate amount
7. Oppose a great deal

3) Opposition to using military force in international conflict (pro-life attitude)

Survey question: How willing should the United States be to use military force to solve international problems? 
1. Extremely willing
2. Very willing
3. Moderately willing
4. A little willing
5. Not at all willing

4) Degree of Conservatism

Survey question: Where would you place yourself on this scale [7PT SCALE LIBERAL-CONSERVATIVE]?
1. Extremely liberal
2. Liberal
3. Slightly liberal
4. Moderate; middle of the road
5. Slightly conservative
6. Conservative
7. Extremely conservative

Since opposition to the death penalty and opposition to using military force in international conflict are both pro-life attitudes, and favoring of abortion rights is a non-pro-life attitude, a significant positive correlation between these variables would support my hypothesis that pro-life attitudes are not consistent between abortion and other issues. Since ____ EDIT EDIT
EDIT
EDIT
EDIT
EDIT
EDIT
EDIT
significant negative correlation between these variables would support my hypothesis that pro-life attitudes are not consistent between abortion and other issues. A significant negative correlation between degree of conservatism and favoring of abortion rights would further support my hypothesis that the pro-life argument for abortion is not based in values, but instead a partisan identification or political argument. However, if these correlations were to be swapped, this would suggest that pro-life attitudes are consistent among many political issues, and that the pro-life abortion argument is a true value, not just political rhetoric.

## Results

```{r}
ggplot(dataset, aes(x = deathpenalty, y = abortion)) +
geom_jitter() + geom_smooth(method = "lm", se = FALSE) +
labs(
x = "Death Penalty Unavorability",
y = "Abortion Reduction Favorability",
title = "Pro-Choice vs Anti-Death Penalty"
)

ggplot(dataset, aes(x = guncontrol, y = abortion)) +
geom_jitter() + geom_smooth(method = "lm", se = FALSE) +
labs(
x = "Assault Rifle Bans Favorability",
y = "Abortion Reduction Favorability",
title = "Pro-Choice vs Pro-Assault Rifles"
)

ggplot(dataset, aes(x = militaryforce, y = abortion)) +
geom_jitter() + geom_smooth(method = "lm", se = FALSE) +
labs(
x = "Military Force Unfavorability",
y = "Abortion Reduction Favorability",
title = "Pro-Choice vs Anti-Military Force"
)

ggplot(dataset, aes(x = ideology, y = abortion)) +
geom_jitter() + geom_smooth(method = "lm", se = FALSE) +
labs(
x = "Conservative Identification",
y = "Abortion Reduction Favorability",
title = "Pro-Choice vs Pro-Assault Rifles"
)

## Regression
regression <- lm(abortion ~ deathpenalty + militaryforce,
                        data = dataset)
summary(regression)

gunsregression <- lm(abortion ~ guncontrol,
                        data = dataset)
summary(gunsregression)

ideologyregression <- lm(abortion ~ ideology, data = dataset)
summary(ideologyregression)
```